---
title: "Universal GSEA with Custom Gene Sets"
author: "Witold Wolski"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Universal GSEA with Custom Gene Sets}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  cache = TRUE
)
```

# Introduction

This vignette demonstrates how to perform **Gene Set Enrichment Analysis (GSEA)** using the universal `GSEA()` function from clusterProfiler. Unlike the built-in functions (`gseGO`, `gseKEGG`), the universal approach allows you to use **any custom gene set collection**.

We will cover:

1. Understanding **TERM2GENE** and **TERM2NAME** data structures
2. Loading gene sets from **MSigDB** using the `msigdbr` package
3. Running **universal GSEA**
4. **Visualization** techniques for GSEA results
5. **Creating custom gene sets** from scratch
6. **Combining custom gene sets** with existing collections

## Required Packages

```{r load-packages}
library(tidyverse)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
library(org.Hs.eg.db)
```

# Understanding TERM2GENE and TERM2NAME

The `GSEA()` function requires gene sets in a specific format defined by two data frames.

## TERM2GENE: The Gene Set Mapping

**TERM2GENE** is a two-column data frame that maps gene set terms to their member genes:

- **Column 1**: Term/pathway identifier (e.g., "HALLMARK_APOPTOSIS")
- **Column 2**: Gene identifier (e.g., "TP53", "BCL2")

Each row represents one gene belonging to one term. A gene can appear in multiple rows if it belongs to multiple gene sets.

```{r term2gene-example}
# Example TERM2GENE structure
example_t2g <- data.frame(
  gs_name = c("PATHWAY_A", "PATHWAY_A", "PATHWAY_A",
              "PATHWAY_B", "PATHWAY_B", "PATHWAY_B"),
  gene_symbol = c("TP53", "BRCA1", "MYC",
                  "EGFR", "KRAS", "TP53")  # Note: TP53 is in both pathways
)

knitr::kable(example_t2g, col.names = c("Term", "Gene"),
             caption = "Example TERM2GENE structure")
```

**Key points:**

- Column **names don't matter** - only the **position** matters (first = term, second = gene)
- One row per gene-term pair (long format)
- Genes can belong to multiple terms
- Gene identifiers must match your ranked gene list

## TERM2NAME: Human-Readable Descriptions

**TERM2NAME** is an optional two-column data frame providing descriptions for each term:

- **Column 1**: Term/pathway identifier (must match TERM2GENE)
- **Column 2**: Human-readable description

```{r term2name-example}
# Example TERM2NAME structure
example_t2n <- data.frame(
  gs_name = c("PATHWAY_A", "PATHWAY_B"),
  gs_description = c("DNA Damage Response Pathway",
                     "Receptor Tyrosine Kinase Signaling")
)

knitr::kable(example_t2n, col.names = c("Term", "Description"),
             caption = "Example TERM2NAME structure")
```

If TERM2NAME is not provided, the term IDs from TERM2GENE will be used as descriptions in the output.

# Loading Gene Sets from MSigDB

The `msigdbr` package provides convenient access to the Molecular Signatures Database (MSigDB).

## Available Collections

```{r msigdb-collections}
# View available collections
msigdbr_collections() |>
  knitr::kable(caption = "MSigDB Collections")
```

## Loading Hallmark Gene Sets

Let's load the Hallmark collection (H) - 50 well-defined biological states and processes.

```{r load-hallmark}
# Load Hallmark gene sets for human
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H")

# Examine the structure
glimpse(hallmark_sets)
```

## Creating TERM2GENE and TERM2NAME from MSigDB

```{r create-t2g-t2n}
# Create TERM2GENE (using gene symbols)
hallmark_t2g <- hallmark_sets |>
  dplyr::select(gs_name, gene_symbol)

# Create TERM2NAME
hallmark_t2n <- hallmark_sets |>
  dplyr::select(gs_name, gs_description) |>
  dplyr::distinct()

# View the results
head(hallmark_t2g, 10)
head(hallmark_t2n, 5)

# Summary
cat("Number of gene sets:", n_distinct(hallmark_t2g$gs_name), "\n")
cat("Total gene-set pairs:", nrow(hallmark_t2g), "\n")
```

# Preparing the Ranked Gene List

GSEA requires a **named numeric vector** of gene-level statistics, sorted in decreasing order.

## Loading Example Data

```{r load-genelist}
# Load example differential expression data
data(geneList, package = "DOSE")

# geneList is a named numeric vector of log2 fold changes
# Names are Entrez IDs
head(geneList)
tail(geneList)

cat("Number of genes:", length(geneList), "\n")
cat("Range:", round(range(geneList), 2), "\n")
```

## Converting Gene Identifiers

The example `geneList` uses Entrez IDs, but our MSigDB gene sets use gene symbols. We need to convert.

```{r convert-ids}
# Convert Entrez IDs to gene symbols
gene_symbols <- AnnotationDbi::mapIds(
  org.Hs.eg.db,
  keys = names(geneList),
  column = "SYMBOL",
  keytype = "ENTREZID",
  multiVals = "first"
)

# Create symbol-based geneList
geneList_symbol <- geneList
names(geneList_symbol) <- gene_symbols

# Remove NAs and re-sort
geneList_symbol <- geneList_symbol[!is.na(names(geneList_symbol))]
geneList_symbol <- sort(geneList_symbol, decreasing = TRUE)

head(geneList_symbol)
cat("Genes after conversion:", length(geneList_symbol), "\n")
```

## Ranking Metrics

Common metrics for creating a ranked gene list:

| Metric | Pros | Cons |
|--------|------|------|
| **log2 Fold Change** | Simple, interpretable | Ignores significance |
| **Moderated t-statistic** | Combines effect size and variance | Less interpretable |
| **Signed -log10(p) × sign(FC)** | Combines significance and direction | Can be extreme |

```{r ranking-metric-example, eval=FALSE}
# Example: Creating a ranked list from DE results
# geneList <- de_results$log2FC
# names(geneList) <- de_results$gene_symbol
# geneList <- sort(geneList, decreasing = TRUE)
```

# Running Universal GSEA

Now we can run GSEA with our custom TERM2GENE and TERM2NAME.

```{r run-gsea}
gsea_result <- GSEA(
  geneList = geneList_symbol,
  TERM2GENE = hallmark_t2g,
  TERM2NAME = hallmark_t2n,
  minGSSize = 10,        # Minimum gene set size

  maxGSSize = 500,       # Maximum gene set size
  pvalueCutoff = 0.05,   # FDR threshold for results
  pAdjustMethod = "BH",  # Benjamini-Hochberg correction
  verbose = FALSE
)

# Number of significant results
cat("Significant gene sets (FDR < 0.05):", nrow(gsea_result), "\n")
```

## Understanding the Output

```{r gsea-output}
# View top results
gsea_result@result |>
  dplyr::select(Description, setSize, NES, pvalue, p.adjust) |>
  head(10) |>
  knitr::kable(digits = c(0, 0, 2, 4, 4),
               caption = "Top 10 GSEA Results")
```

**Output columns explained:**

| Column | Description |
|--------|-------------|
| `ID` | Term identifier from TERM2GENE |
| `Description` | Term description from TERM2NAME |
| `setSize` | Number of genes in the set (after filtering) |
| `enrichmentScore` | Raw enrichment score (ES) |
| `NES` | Normalized enrichment score (comparable across sets) |
| `pvalue` | Nominal p-value |
| `p.adjust` | FDR-adjusted p-value |
| `qvalue` | Q-value (another FDR measure) |
| `rank` | Position where ES reaches maximum |
| `leading_edge` | Statistics about leading edge genes |
| `core_enrichment` | Genes in the leading edge (driving the enrichment) |

# GSEA Visualizations

The `enrichplot` package provides several visualization options for GSEA results.

## Dot Plot

The dot plot shows enriched terms with gene count and adjusted p-value.

```{r dotplot, fig.height=8}
dotplot(gsea_result,
        showCategory = 20,
        title = "GSEA: Hallmark Gene Sets",
        font.size = 10)
```

The x-axis shows Gene Ratio (proportion of leading edge genes), dot size represents gene set size, and color indicates adjusted p-value.

## Ridge Plot

Ridge plots show the distribution of fold changes for genes in each enriched set.

```{r ridgeplot, fig.height=8}
ridgeplot(gsea_result, showCategory = 15) +
  labs(x = "Enrichment Distribution",
       title = "GSEA Ridge Plot") +
  theme(axis.text.y = element_text(size = 9))
```

This visualization helps interpret whether pathways are up- or down-regulated based on where the density is centered relative to zero.

## GSEA Plot (Running Enrichment Score)

The classic GSEA visualization showing the running enrichment score along the ranked list.

### Single Gene Set

```{r gseaplot-single, fig.height=5}
gseaplot2(gsea_result,
          geneSetID = 1,
          title = gsea_result$Description[1])
```

**Interpretation:**

- **Top panel**: Running enrichment score - peaks indicate where gene set members are concentrated
- **Middle panel**: Locations of gene set members in the ranked list (tick marks)
- **Bottom panel**: The ranked metric values

### Multiple Gene Sets

Compare enrichment profiles across several pathways:

```{r gseaplot-multi, fig.height=6}
gseaplot2(gsea_result,
          geneSetID = 1:4,
          pvalue_table = TRUE,
          ES_geom = "line")
```

## Enrichment Map (emapplot)

The enrichment map clusters similar gene sets based on overlapping genes.

```{r emapplot, fig.height=7}
# First, calculate pairwise term similarity
gsea_sim <- pairwise_termsim(gsea_result)

# Create the enrichment map
emapplot(gsea_sim,
         showCategory = 20) +
  labs(title = "Enrichment Map: Gene Set Similarity Network")
```

Connected nodes share genes, helping identify functional clusters.

## Gene-Concept Network (cnetplot)

Shows which genes contribute to enriched terms, colored by fold change.

```{r cnetplot, fig.height=8, fig.width=10}
cnetplot(gsea_result,
         showCategory = 5,
         foldChange = geneList_symbol,
         cex_label_gene = 0.6,
         cex_label_category = 0.9) +
  labs(title = "Gene-Concept Network")
```

This reveals genes that participate in multiple enriched pathways (hubs).

## Heatplot

A heatmap representation of gene-concept relationships.

```{r heatplot, fig.height=5, fig.width=12}
heatplot(gsea_result,
         showCategory = 6,
         foldChange = geneList_symbol) +
  labs(title = "Gene-Pathway Heatmap")
```

# Creating Custom Gene Sets from Scratch

You can create your own gene sets based on domain knowledge, literature, or experimental data.

## Single Custom Gene Set

```{r custom-single}
# Example: Create a custom "DNA Repair" gene set
my_dna_repair_t2g <- data.frame(
  gs_name = "CUSTOM_DNA_REPAIR",
  gene_symbol = c("TP53", "BRCA1", "BRCA2", "ATM", "ATR",
                  "CHEK1", "CHEK2", "RAD51", "XRCC1", "PARP1",
                  "MLH1", "MSH2", "MSH6", "NBN", "MRE11",
                  "FANCD2", "FANCA", "PALB2", "BLM", "WRN")
)

my_dna_repair_t2n <- data.frame(
  gs_name = "CUSTOM_DNA_REPAIR",
  gs_description = "Custom DNA Repair Genes (manual curation)"
)

knitr::kable(my_dna_repair_t2g,
             caption = "Custom DNA Repair Gene Set")
```

## Multiple Custom Gene Sets

```{r custom-multiple}
# Create multiple custom gene sets
my_custom_t2g <- data.frame(
  gs_name = c(
    # Apoptosis regulators
    rep("CUSTOM_APOPTOSIS", 10),
    # Cell cycle checkpoints
    rep("CUSTOM_CELL_CYCLE", 8),
    # Metabolic enzymes
    rep("CUSTOM_METABOLISM", 7)
  ),
  gene_symbol = c(
    # Apoptosis
    "BCL2", "BAX", "CASP3", "CASP8", "CASP9",
    "CYCS", "APAF1", "BID", "XIAP", "MCL1",
    # Cell cycle
    "CDK1", "CDK2", "CDK4", "CCND1", "CCNE1",
    "RB1", "E2F1", "CDKN1A",
    # Metabolism
    "HK1", "HK2", "PFKM", "PKM", "LDHA", "G6PD", "IDH1"
  )
)

my_custom_t2n <- data.frame(
  gs_name = c("CUSTOM_APOPTOSIS", "CUSTOM_CELL_CYCLE", "CUSTOM_METABOLISM"),
  gs_description = c(
    "Custom Apoptosis Pathway Genes",
    "Custom Cell Cycle Checkpoint Genes",
    "Custom Metabolic Enzyme Genes"
  )
)

# Summary
my_custom_t2g |>
  group_by(gs_name) |>
  summarise(n_genes = n(), .groups = "drop") |>
  knitr::kable(caption = "Custom Gene Sets Summary")
```

# Combining Custom Gene Sets with Existing Collections

A powerful feature is adding your custom gene sets to established collections like MSigDB.

## Using rbind() to Combine

```{r combine-genesets}
# Combine Hallmark sets with custom sets
combined_t2g <- rbind(
  hallmark_t2g,
  my_custom_t2g
)

combined_t2n <- rbind(
  hallmark_t2n,
  my_custom_t2n
)

# Verify custom sets are included
combined_t2g |>
  dplyr::group_by(gs_name) |>
  dplyr::summarise(n_genes = n(), .groups = "drop") |>
  dplyr::filter(str_detect(gs_name, "^CUSTOM")) |>
  knitr::kable(caption = "Custom Sets in Combined Collection")

cat("Total gene sets:", n_distinct(combined_t2g$gs_name), "\n")
cat("Total gene-term pairs:", nrow(combined_t2g), "\n")
```

## Running GSEA with Combined Sets

```{r gsea-combined}
gsea_combined <- GSEA(
  geneList = geneList_symbol,
  TERM2GENE = combined_t2g,
  TERM2NAME = combined_t2n,
  minGSSize = 5,         # Lower threshold to include smaller custom sets
  maxGSSize = 500,
  pvalueCutoff = 1,      # Keep all results for inspection
  verbose = FALSE
)

# Check results for custom gene sets
gsea_combined@result |>
  dplyr::filter(str_detect(ID, "^CUSTOM")) |>
  dplyr::select(Description, setSize, NES, pvalue, p.adjust) |>
  knitr::kable(digits = c(0, 0, 2, 4, 4),
               caption = "GSEA Results for Custom Gene Sets")
```

## Visualizing Custom Sets in Context

```{r custom-viz}
# Find the indices of custom sets
custom_indices <- which(str_detect(gsea_combined@result$ID, "^CUSTOM"))

if (length(custom_indices) > 0) {
  gseaplot2(gsea_combined,
            geneSetID = custom_indices,
            pvalue_table = TRUE,
            title = "Custom Gene Sets: Enrichment Profiles")
}
```

# Working with GMT Files

GMT (Gene Matrix Transposed) is a standard format for gene sets.

## GMT File Format

```
PATHWAY_NAME<TAB>DESCRIPTION<TAB>GENE1<TAB>GENE2<TAB>GENE3<TAB>...
```

Each line represents one gene set:
- Field 1: Gene set name
- Field 2: Description (can be "NA" or URL)
- Field 3+: Gene identifiers

## Reading GMT Files

```{r read-gmt, eval=FALSE}
# Read a GMT file (example - file doesn't exist)
my_gmt <- clusterProfiler::read.gmt("path/to/geneset.gmt")

# The result is a TERM2GENE data frame
head(my_gmt)
```

## Writing GMT Files

Here's a function to export your custom gene sets as GMT:

```{r write-gmt-function}
#' Write TERM2GENE and TERM2NAME to GMT format
#'
#' @param t2g TERM2GENE data frame (two columns: term, gene)
#' @param t2n TERM2NAME data frame (two columns: term, description). Optional.
#' @param file Output file path
write_gmt <- function(t2g, t2n = NULL, file) {
  terms <- unique(t2g[[1]])

  lines <- sapply(terms, function(term) {
    genes <- t2g[t2g[[1]] == term, 2]
    desc <- if (!is.null(t2n)) {
      idx <- which(t2n[[1]] == term)
      if (length(idx) > 0) t2n[[2]][idx[1]] else "NA"
    } else {
      "NA"
    }
    paste(c(term, desc, genes), collapse = "\t")
  })

  writeLines(lines, file)
  cat("Wrote", length(lines), "gene sets to", file, "\n")
}

# Example usage (not run)
# write_gmt(my_custom_t2g, my_custom_t2n, "my_custom_sets.gmt")
```

# Other MSigDB Collections

## Loading Different Collections

```{r other-collections, eval=FALSE}
# C2: Curated gene sets (includes KEGG, Reactome, BioCarta)
c2_kegg <- msigdbr(species = "Homo sapiens",
                   category = "C2",
                   subcategory = "CP:KEGG") |>
  select(gs_name, gene_symbol)

c2_reactome <- msigdbr(species = "Homo sapiens",
                       category = "C2",
                       subcategory = "CP:REACTOME") |>
  select(gs_name, gene_symbol)

# C5: GO gene sets
c5_bp <- msigdbr(species = "Homo sapiens",
                 category = "C5",
                 subcategory = "GO:BP") |>
  select(gs_name, gene_symbol)

# C7: Immunologic signatures
c7_imm <- msigdbr(species = "Homo sapiens", category = "C7") |>
  select(gs_name, gene_symbol)

# C8: Cell type signatures
c8_cell <- msigdbr(species = "Homo sapiens", category = "C8") |>
  select(gs_name, gene_symbol)
```

## Working with Mouse Data

```{r mouse-example}
# MSigDB provides ortholog-mapped gene sets for many species
msigdbr_species() |>
  head(10) |>
  knitr::kable(caption = "Available Species in msigdbr")
```

```{r mouse-hallmark, eval=FALSE}
# Load Hallmark for mouse
mouse_hallmark <- msigdbr(species = "Mus musculus", category = "H") |>
  select(gs_name, gene_symbol)
```

# Summary

## Key Concepts

**TERM2GENE:**

- Two-column data frame (term, gene)
- One row per gene-term membership
- Column positions matter, names don't
- Gene IDs must match your ranked list

**TERM2NAME:**

- Two-column data frame (term, description)
- One row per term
- Optional but recommended

**Custom Gene Sets:**

- Create simple data frames
- Use `rbind()` to combine with existing collections
- Adjust `minGSSize` for small custom sets
- Export to GMT for sharing

## Visualization Summary

| Function | Purpose |
|----------|---------|
| `dotplot()` | Overview of all enriched terms |
| `ridgeplot()` | Fold change distributions per term |
| `gseaplot2()` | Running enrichment score |
| `emapplot()` | Term similarity network |
| `cnetplot()` | Gene-term network |
| `heatplot()` | Gene-term heatmap |

# Session Information

```{r session-info}
sessionInfo()
```

# References

- Yu G, Wang L, Han Y, He Q (2012). clusterProfiler: an R package for comparing biological themes among gene clusters. OMICS: A Journal of Integrative Biology, 16(5):284-287.
- Subramanian A, Tamayo P, et al. (2005). Gene set enrichment analysis: A knowledge-based approach. PNAS 102(43):15545-15550.
- Liberzon A, et al. (2015). The Molecular Signatures Database Hallmark Gene Set Collection. Cell Systems 1(6):417-425.

**Useful Links:**

- clusterProfiler book: https://yulab-smu.top/biomedical-knowledge-mining-book/
- MSigDB: https://www.gsea-msigdb.org/gsea/msigdb/
- msigdbr package: https://cran.r-project.org/package=msigdbr
- enrichplot visualizations: https://yulab-smu.top/biomedical-knowledge-mining-book/enrichplot.html
